En Garde! Unguarded Iteration for Reversible Computation in the Delay Monad\thanks{Niccol\`o Veltri was supported by a research grant (13156) from VILLUM FONDEN.}

### 2019-05-06 mpc19/paper.tex
overview + term interpretation

% Semantics
\newcommand{\semTy}[1]{{\llbracket #1 \rrbracket}_\mathsf{Ty}}
\newcommand{\semMu}[1]{{\llbracket #1 \rrbracket}_\mu}
\newcommand{\semTm}[1]{{\llbracket #1 \rrbracket}_{\lr}}
\newcommand{\semEq}[1]{{\llbracket #1 \rrbracket}_{\LR}}
\newcommand{\semTrace}{\mathsf{trace}_{\D\simeq}}

@@

The full \Pio\ calculus cannot be modelled in the same framework of
\cite{CaretteS16}, since Martin-L\"of type theory is a total language
which in particolar disallows the specification of a trace
operator on types. Consequently, it is necessary to move to a setting
supporting the existence of partial maps, and in type theory this can
be done by using monads, by considering partiality as an effect. Our
choice fell on the coinductive delay monad, introduced by Capretta
\cite{Capretta05} as a way of representing general recursive functions
in Martin-L\"of type theory. The delay datatype has been employed in a
large number of applications, ranging from operational semantics of
functional languages \cite{Danielsson12} to formalization of domain
theory in type theory \cite{BentonKV09} and normalization by
evaluation \cite{AbelC14}. Here it is used for giving denotational
semantics to \Pio. In particular, we show how to endow the delay
datatype with a trace combinator, whose construction factors
through the specification of a uniform iteration operator
\cite{GoncharovSRJ18,GoncharovMR16}.

@@

@@

In this section, we present the syntax of \Pio. The 1-structure of
\Pio, i.e. its types and terms, has originaly been introduced by James
and Sabry \cite{JamesS12}. In particular, we include the presence of
recursive types and a primitive trace combinator.  Following Carette
and Sabry's formalization of the finitary fragment of \Pio, we
consider a collection of equivalences between terms. Our list of
axioms notably differs from theirs in that we do not require each term to be
a total isomorphism, we ask only for the existence of a partial
inverse.

@@

\begin{figure}
\[
\arraycolsep=8pt\def\arraystretch{2.8}
\begin{array}{ccc}
%\hline 
\infer{\Z : \Ty\,n}{} 
& \infer{\I : \Ty\,n}{}
& \infer{\Var\,i : \Ty\,n}{i : \mathsf{Fin}\,n}  \\
\infer{A \oplus B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
& \infer{A \otimes B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
& \infer{\mu A : \Ty\,n}{A : \Ty\,(n+1)} 
%\\ \hline
\end{array}
\]
\caption{Types of \Pio, as formalized in Agda}
\label{fig:types}
\end{figure}

@@

\begin{figure}
\[
\arraycolsep=7pt\def\arraystretch{1.8}
\begin{array}{cc}%{|cc|}
%\hline 
\id : A \lr A
& \infer{g \bullet f : A \lr C}{g : B \lr C & f : A \lr B} \\ [.25cm]
\infer{f \oplus g : A \oplus B \lr C \oplus D}{f : A \lr C & g : B \lr D} 
& \infer{f \otimes g : A \otimes B \lr C \otimes D}{f : A \lr C & g : B \lr D} \\
\lambda_{\oplus} : \Z \oplus A \lr A
& \lambda_{\oplus}^{-1} : A \lr \Z \oplus A\\
\lambda_{\otimes} : \I \otimes A \lr A
& \lambda_{\otimes}^{-1} : A \lr \I \otimes A\\
\alpha_{\oplus} : (A \oplus B) \oplus C \lr A \oplus (B \oplus C)
& \alpha_{\oplus}^{-1} : A \oplus (B \oplus C) \lr (A \oplus B) \oplus C \\
\alpha_{\otimes} : (A \otimes B) \otimes C \lr A \otimes (B \otimes C)
& \alpha_{\otimes}^{-1} : A \otimes (B \otimes C) \lr (A \otimes B) \otimes C \\
\sigma_{\oplus} : A \oplus B \lr B \oplus A
& \sigma_{\otimes} : A \otimes B \lr B \otimes A \\
\kappa : \Z \otimes A \lr \Z 
& \delta : (A \oplus B) \otimes C \lr (A \otimes C) \oplus (B \otimes C) \\
\kappa^{-1} : \Z \lr \Z \otimes A
& \delta^{-1} : (A \otimes C) \oplus (B \otimes C)  \lr (A \oplus B) \otimes C \\
\fold : A[\mu X.A/X] \lr \mu X.A
& \unfold : \mu X.A \lr A[\mu X.A/X] \\[.25cm]
\multicolumn{2}{c}{\infer{\trace \,f: A \lr B}{f : A \oplus C \lr B \oplus C}} \\
%\hline
\end{array}
\]
\caption{Terms of \Pio}
\label{fig:programs}
\end{figure}

@@

\begin{figure}
\[
\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{cc} %{|cc|} \hline 
\multicolumn{2}{c}{\mathsf{naturality_L} : f \bullet \trace\,g \LR \trace\,((f \oplus
  \id) \bullet g)} \\ 
\multicolumn{2}{c}{\mathsf{naturality_R} : \trace\,g \bullet f \LR \trace\,(g \bullet (f 
  \oplus \id))} \\  
\multicolumn{2}{c}{\mathsf{dinaturality} : \trace \,((\id \oplus f) \bullet g) \LR \trace
  \,(g \bullet (\id \oplus f))} \\
\multicolumn{2}{c}{\mathsf{superposing} : \trace \,(\alpha_\oplus^{-1}
  \bullet (\id \oplus f) \bullet \alpha_\oplus) \LR \id \oplus \trace \,f} \\ 
\multicolumn{2}{c}{\mathsf{vanishing_\oplus} : \trace \,f \LR \trace \,(\trace
  \,(\alpha_\oplus^{-1} \bullet f \bullet \alpha_\oplus))} \\ 
\mathsf{vanishing_\Z} : f \LR \rho_\oplus^{-1} \bullet \trace \,f
  \bullet \rho_\oplus 
&
\mathsf{yanking} : \trace\, \sigma_\oplus \LR \id
 \\
\multicolumn{2}{c}{\mathsf{tracePIso} : \trace\,f \bullet \trace\,(\dagger\,f)
  \bullet \trace\,f \LR \trace \,f} \\
%%  \\[.25cm]
%% \multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
%%   \bullet \trace\,f \LR \trace \,f}{p : f \bullet \dagger\,f \bullet
%%   f \LR f}} \\
\mathsf{foldIso} : \fold \bullet \unfold \LR \id 
&
\mathsf{unfoldIso} : \unfold \bullet \fold \LR \id
\\
\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet \dagger \,f
  \bullet g \bullet \dagger \, g \LR g \bullet \dagger \,g
  \bullet f \bullet \dagger \, f}
%\\ \hline
\end{array}
\]
\caption{Selection of term equivalences of \Pio}
\label{fig:programequivs}
\end{figure}

@@

In $\Dapprox$, our definition of partial isomorphisms is equivalent to
the standard categorical one~\cite{Kastl79} (see also \cite{CockettL02}),
which, translated in our type-theoretical setting, is
\begin{align*}
\pisoalt \,f = (g : B \to \Delay A) \times f \diamond g \diamond
  f \approx f \times g \diamond f \diamond g \approx g 
\end{align*}
We denote $A \simeq B$ the type of partial isomorphisms between
$A$ and $B$:
\[
A \simeq B = (f : A \to \Delay B) \times \piso \,f
\]
We call $\Inv \Dapprox$ the subcategory of $\Dapprox$ consisting of
(equivalence classes of) partial isomorphisms. Note that $\Inv\Dapprox$
inherits neither partial products nor coproducts of $\Dapprox$, as the
universal mapping property fails in both cases. However, it can be shown that
in the category $\Inv\Dapprox$, $0$ is a zero object, $A + B$ is the
disjointness tensor product of $A$ and $B$ (in the sense of Giles~\cite{Giles})
with unit $0$, and $A \times B$ a monoidal product of $A$ and $B$ with unit $1$
(though it is \emph{not} an inverse product in the sense of
Giles~\cite{Giles}, as that would imply decidable equality on all objects).

@@

An Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
whose Kleilsi category supports unguarded uniform iteration. More
precisely\footnote{Here we give the definition of Elgot monad on
  $\Set$, but the definition of Elgot monad makes sense in any
  category with binary coproducts.}, a monad $\T$ is Elgot if there
exists an operation
\[
\iter_\T : (A \to \T \,(B + A)) \to A \to \T\,B
\]
satisfying the following axioms:
\[\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{c}
\mathsf{fixpoint} :
\iter_\T\,f \equiv \copair {\eta_\T}{\iter_\T \,f} \diamond_\T f
\\
\mathsf{naturality} : g \diamond_\T \iter_\T\,f \equiv
  \iter_\T\,(\copair{\T \inl \comp g} {\eta \comp \inr} \diamond_\T f)
\\
\mathsf{codiagonal} :
\iter_\T \,(\iter_\T\,g) \equiv \iter_\T\,(\T \copair {\id} {\inr}
  \comp g)
\\[.25cm]
\infer{\mathsf{uniformity}\,p : \iter_\T \,f \comp h \equiv \iter_\T
  \,g}{p : f \comp h \equiv \T (\id + h) \comp g}
\end{array}
\]
where $\eta_\T$ is the unit of $\T$ and $\diamond_\T$ denotes morphism
composition in the Kleisli category of $\T$.
The standard definition of uniform iteration operator includes the
dinaturality axiom, which has recently been discovered to be derivable
from the other laws \cite{EsikG16,GoncharovMR16}.

The delay monad is an Elgot monad for which the axioms holds up to
weak bisimilarity, not propositional equality. In other words, the
category $\Dapprox$ can be endowed with a uniform iteration
operator. The specification of the iteration operator relies on an
auxiliary function $\iterD'$ corecursively defined as follows:
\[ %\arraycolsep=8pt
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\iterD' : (A \to \Delay (B + A)) \to \Delay (B + A) \to \Delay B} \\
\iterD' \,f \,(\now \,(\inl\, b)) &= \now \,b \\
\iterD' \,f \,(\now \,(\inr\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
\iterD' \,f \,(\later \,x) &= \later \,(\iterD' \,f \,x) \\
 \\
\multicolumn{2}{l}{\iterD : (A \to \Delay (B + A)) \to A \to \Delay B} \\
\multicolumn{2}{l}{\iterD \,f \,a = \iterD' \,f \,(f \,a)}
\end{array}
\]

%% \begin{align*}
%% & \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
%% & \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
%% & \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
%% & \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
%% & \\
%% & \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
%% & \iterD \,f \,a = \iterD' \,f \,(f \,a)
%% \end{align*}
The definition above can be given the following intuitive
explaination. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
not terminate as well. If $f\,a$ terminates, there are two
possibilities: either $f\,a$ converges to $\inr\,b$, in which case
$\iterD\,f\,a$ terminates returning the value $b$; or $f\,a$ converges
to $\inl\,a'$, in which case we repeat the precedure by replacing $a$
with $a'$. Notice that in the latter case we also add one occurrence
of $\later$ to the total computation time. This addition is necessary
for ensuring the productivity of the corecursively defined function
$\iterD'$. In fact, by changing the second line of its specification to
$\iterD' \,f \,(\now \,(\inr\, a)) = \iterD' \,f \,(f a)$
and taking $f = \inrD$, we would have that $\iterD' \,f \,(\now
\,(\inr\, a))$ unfolds indefinitely without producing any output. In
Agda, such a definition would be rightfully rejected by the termination
checker.

@@

Terms of \Pio\ are modelled as morphism of $\Inv\Dapprox$,
i.e. partial isomorphisms. Formally, we define an operation:
\[
\semTm - : (A \lr B) \to \semTy A \, \rho \simeq \semTy B \, \rho
\]
We refer the interested reader to our Agda formalization for a
complete definition of the interpretation of terms. Here we limit
ourselves to emphasize
the semantics of the trace operator: $\semTm {\trace\,f} =
\semTrace\,\semTm f$.

Term equivalences of \Pio\ are modelled as morphism equalities in
$\Inv\Dapprox$, i.e. proofs of weak bisimilarity between two
morpshisms. Formally, we define an operation:
\[
\semEq - : (f \LR g) \to \semTm f \approx \semTm g
\]
Again we refer the interested reader to our Agda formalization for a
complete definition of the interpretation of term equivalences.

\section{Conclusions}
In this paper, we have extended the work of Carette and Sabry~\cite{CaretteS16}
to a (fully formalized) two-level calculus of \Pio{} programs and program
equivalences. Key in this effort was the use of the Kleisli category of the
delay monad on $\Set$ under weak bisimilarity, which turned out to support
iteration via a trace that preserves all partial isomorphisms, in this way
giving semantics to the dagger trace of \Pio. Further, the work was formalized
using Agda 2.6.0.

It is natural to wonder if our work can be ported to other monads of partiality
in Martin-L\"of type theory. As already discussed in the end of Section
\ref{sec:trace}, the maybe monad is not suitable for modelling a
well-behaved trace combinator without the assumption of classical principles
such as \textsf{LPO}. The partial map classifier \cite{UustaluV17-2,EscardoK17}
$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where $\mathsf{Prop}$ is
the type of propositions (types with at most one inhabitant), supports the
existence of a uniform iteration operator and therefore a trace. Nevertheless,
the specification of iteration is more complicated then the one presented in
Section \ref{sec:elgot} for the delay monad, which is a simple corecursive
definition. The Elgot monad structure of $\mathsf{PMC}$ follows from its
Kleisli category being a join restriction category, so iteration is defined in
terms of least upper bounds of certain chains of morphisms. The subcategory of
partial isomorphisms of the Kleisli category of $\mathsf{PMC}$ supports a
dagger trace combinator, which can be proved following the general strategy in
\cite{KaarsgaardAG17}. The exact same observations apply to the partiality
monad in homotopy type theory \cite{AltenkirchDK17,ChapmanUV19}, to which the
quotiented delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
isomorphic under the assumption of countable choice.

>>>>

% Semantics
\newcommand{\semTy}[1]{{\llbracket #1 \rrbracket}_\mathsf{Ty}}
\newcommand{\semMu}[1]{{\llbracket #1 \rrbracket}_\mu}
\newcommand{\semTm}[1]{{\llbracket #1 \rrbracket}_{\lr}}
\newcommand{\semEq}[1]{{\llbracket #1 \rrbracket}_{\!\!\LR}}
\newcommand{\semTrace}{\mathsf{trace}_{\D\simeq}}

@@

The full \Pio\ calculus cannot be modelled in the same framework of
\cite{CaretteS16}, since Martin-L\"of type theory is a total language
which in particular disallows the specification of a trace
operator on types. Consequently, it is necessary to move to a setting
supporting the existence of partial maps, and in type theory this can
be done by using monads, by considering partiality as an effect. Our
choice fell on the coinductive delay monad, introduced by Capretta
\cite{Capretta05} as a way of representing general recursive functions
in Martin-L\"of type theory. The delay datatype has been employed in a
large number of applications, ranging from operational semantics of
functional languages \cite{Danielsson12} to formalization of domain
theory in type theory \cite{BentonKV09} and normalization by
evaluation \cite{AbelC14}. Here it is used for giving denotational
semantics to \Pio. In particular, we show how to endow the delay
datatype with a trace combinator, whose construction factors
through the specification of a uniform iteration operator
\cite{GoncharovSRJ18,GoncharovMR16}.

@@

\paragraph{Overview}
In Section \ref{sec:syntax}, we present the syntax of \Pio\ as
formalized in Agda, with particular emphasis on recursive types and
the trace operator. In Section \ref{sec:delay}, we recall the
definition of Capretta's delay datatype and weak bisimilarity. We
discuss finite products and coproducts in the Kleisli category of the
delay monad and we introduce the category of partial isomorphisms that
serves as the denotational model of \Pio. In Section
\ref{sec:elgot}, we build a complete Elgot monad structure on the delay
datatype, that allows the encoding of a dagger trace operator in the
category of partial isomorphisms. In Section \ref{sec:interpretation},
we formally describe the interpretation of \Pio\ types, terms and
terms equivalences. We conclude in Section \ref{sec:conclusions} with
some final remarks and discussion on future work.

@@

In this section, we present the syntax of \Pio. The 1-structure of
\Pio, i.e. its types and terms, has originally been introduced by James
and Sabry \cite{JamesS12}. In particular, we include the presence of
recursive types and a primitive trace combinator.  Following Carette
and Sabry's formalization of the finitary fragment of \Pio, we
consider a collection of equivalences between terms. Our list of
axioms notably differs from theirs in that we do not require each term to be
a total isomorphism, we ask only for the existence of a partial
inverse.

@@

\begin{figure}[h]
\[
\arraycolsep=8pt\def\arraystretch{2.8}
\begin{array}{ccc}
%\hline 
\infer{\Z : \Ty\,n}{} 
& \infer{\I : \Ty\,n}{}
& \infer{\Var\,i : \Ty\,n}{i : \mathsf{Fin}\,n}  \\
\infer{A \oplus B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
& \infer{A \otimes B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
& \infer{\mu A : \Ty\,n}{A : \Ty\,(n+1)} 
%\\ \hline
\end{array}
\]
\caption{Types of \Pio, as formalized in Agda}
\label{fig:types}
\end{figure}

@@

\begin{figure}[t]
\[
\arraycolsep=7pt\def\arraystretch{1.8}
\begin{array}{cc}%{|cc|}
%\hline 
\id : A \lr A
& \infer{g \bullet f : A \lr C}{g : B \lr C & f : A \lr B} \\ [.25cm]
\infer{f \oplus g : A \oplus B \lr C \oplus D}{f : A \lr C & g : B \lr D} 
& \infer{f \otimes g : A \otimes B \lr C \otimes D}{f : A \lr C & g : B \lr D} \\
\lambda_{\oplus} : \Z \oplus A \lr A
& \lambda_{\oplus}^{-1} : A \lr \Z \oplus A\\
\lambda_{\otimes} : \I \otimes A \lr A
& \lambda_{\otimes}^{-1} : A \lr \I \otimes A\\
\alpha_{\oplus} : (A \oplus B) \oplus C \lr A \oplus (B \oplus C)
& \alpha_{\oplus}^{-1} : A \oplus (B \oplus C) \lr (A \oplus B) \oplus C \\
\alpha_{\otimes} : (A \otimes B) \otimes C \lr A \otimes (B \otimes C)
& \alpha_{\otimes}^{-1} : A \otimes (B \otimes C) \lr (A \otimes B) \otimes C \\
\sigma_{\oplus} : A \oplus B \lr B \oplus A
& \sigma_{\otimes} : A \otimes B \lr B \otimes A \\
\kappa : \Z \otimes A \lr \Z 
& \delta : (A \oplus B) \otimes C \lr (A \otimes C) \oplus (B \otimes C) \\
\kappa^{-1} : \Z \lr \Z \otimes A
& \delta^{-1} : (A \otimes C) \oplus (B \otimes C)  \lr (A \oplus B) \otimes C \\
\fold : A[\mu X.A/X] \lr \mu X.A
& \unfold : \mu X.A \lr A[\mu X.A/X] \\[.25cm]
\multicolumn{2}{c}{\infer{\trace \,f: A \lr B}{f : A \oplus C \lr B \oplus C}} \\
%\hline
\end{array}
\]
\caption{Terms of \Pio}
\label{fig:programs}
\end{figure}

@@

\begin{figure}[t]
\[
\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{cc} %{|cc|} \hline 
\multicolumn{2}{c}{\mathsf{naturality_L} : f \bullet \trace\,g \LR \trace\,((f \oplus
  \id) \bullet g)} \\ 
\multicolumn{2}{c}{\mathsf{naturality_R} : \trace\,g \bullet f \LR \trace\,(g \bullet (f 
  \oplus \id))} \\  
\multicolumn{2}{c}{\mathsf{dinaturality} : \trace \,((\id \oplus f) \bullet g) \LR \trace
  \,(g \bullet (\id \oplus f))} \\
\multicolumn{2}{c}{\mathsf{superposing} : \trace \,(\alpha_\oplus^{-1}
  \bullet (\id \oplus f) \bullet \alpha_\oplus) \LR \id \oplus \trace \,f} \\ 
\multicolumn{2}{c}{\mathsf{vanishing_\oplus} : \trace \,f \LR \trace \,(\trace
  \,(\alpha_\oplus^{-1} \bullet f \bullet \alpha_\oplus))} \\ 
\mathsf{vanishing_\Z} : f \LR \rho_\oplus^{-1} \bullet \trace \,f
  \bullet \rho_\oplus 
&
\mathsf{yanking} : \trace\, \sigma_\oplus \LR \id
 \\
\multicolumn{2}{c}{\mathsf{tracePIso} : \trace\,f \bullet \trace\,(\dagger\,f)
  \bullet \trace\,f \LR \trace \,f} \\
%%  \\[.25cm]
%% \multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
%%   \bullet \trace\,f \LR \trace \,f}{p : f \bullet \dagger\,f \bullet
%%   f \LR f}} \\
\mathsf{foldIso} : \fold \bullet \unfold \LR \id 
&
\mathsf{unfoldIso} : \unfold \bullet \fold \LR \id
\\
\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet \dagger \,f
  \bullet g \bullet \dagger \, g \LR g \bullet \dagger \,g
  \bullet f \bullet \dagger \, f}
%\\ \hline
\end{array}
\]
\caption{Selection of term equivalences of \Pio}
\label{fig:programequivs}
\end{figure}

@@

In $\Dapprox$, our definition of partial isomorphisms is equivalent to
the standard categorical one~\cite{Kastl79} (see also \cite{CockettL02}),
which, translated in our type-theoretical setting, is
\begin{align*}
\pisoalt \,f = (g : B \to \Delay A) \times f \diamond g \diamond
  f \approx f \times g \diamond f \diamond g \approx g 
\end{align*}
We denote $A \simeq B$ the type of partial isomorphisms between
$A$ and $B$:
\[
A \simeq B = (f : A \to \Delay B) \times \piso \,f
\]
We call $\Inv \Dapprox$ the subcategory of $\Dapprox$ consisting of
(equivalence classes of) partial isomorphisms. Note that
$\Inv\Dapprox$ inherits neither partial products nor coproducts of
$\Dapprox$, as the universal mapping property fails in both
cases. However, it can be shown that in the category $\Inv\Dapprox$,
$0$ is a zero object, $A + B$ is the disjointness tensor product of
$A$ and $B$ (in the sense of Giles~\cite{Giles}) with unit $0$, and
$A \times B$ a monoidal product of $A$ and $B$ with unit $1$ (though
it is \emph{not} an inverse product in the sense of
Giles~\cite{Giles}, as that would imply decidable equality on all
objects).  In particular, we can derive the following operations,
modelling the \Pio\ term constructors $\oplus$ and $\otimes$:
\begin{align*}
& \times_{\D \simeq} : A \simeq C \to B \simeq D \to A \times B
  \simeq C \times D \\
& +_{\D \simeq} : A \simeq C \to B \simeq D \to A + B
  \simeq C + D
\end{align*}

@@

A complete Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
whose Kleisli category supports unguarded uniform iteration. More
precisely\footnote{Here we give the definition of complete Elgot monad on
  $\Set$, but the definition of complete Elgot monad makes sense in any
  category with finite coproducts.}, a monad $\T$ is Elgot if there
exists an operation
\[
\iter_\T : (A \to \T \,(B + A)) \to A \to \T\,B
\]
satisfying the following axioms:
\[\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{c}
\mathsf{fixpoint} :
\iter_\T\,f \equiv \copair {\eta_\T}{\iter_\T \,f} \diamond_\T f
\\
\mathsf{naturality} : g \diamond_\T \iter_\T\,f \equiv
  \iter_\T\,(\copair{\T \inl \comp g} {\eta \comp \inr} \diamond_\T f)
\\
\mathsf{codiagonal} :
\iter_\T \,(\iter_\T\,g) \equiv \iter_\T\,(\T \copair {\id} {\inr}
  \comp g)
\\[.25cm]
\infer{\mathsf{uniformity}\,p : \iter_\T \,f \comp h \equiv \iter_\T
  \,g}{p : f \comp h \equiv \T (\id + h) \comp g}
\end{array}
\]
where $\eta_\T$ is the unit of $\T$ and $\diamond_\T$ denotes morphism
composition in the Kleisli category of $\T$.
The standard definition of uniform iteration operator includes the
dinaturality axiom, which has recently been discovered to be derivable
from the other laws \cite{EsikG16,GoncharovMR16}.

The delay monad is a complete Elgot monad for which the axioms holds up to
weak bisimilarity, not propositional equality. In other words, the
category $\Dapprox$ can be endowed with a uniform iteration
operator. The specification of the iteration operator relies on an
auxiliary function $\iterD'$ corecursively defined as follows:
\[ %\arraycolsep=8pt
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\iterD' : (A \to \Delay (B + A)) \to \Delay (B + A) \to \Delay B} \\
\iterD' \,f \,(\now \,(\inl\, b)) &= \now \,b \\
\iterD' \,f \,(\now \,(\inr\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
\iterD' \,f \,(\later \,x) &= \later \,(\iterD' \,f \,x) \\
 \\
\multicolumn{2}{l}{\iterD : (A \to \Delay (B + A)) \to A \to \Delay B} \\
\multicolumn{2}{l}{\iterD \,f \,a = \iterD' \,f \,(f \,a)}
\end{array}
\]

%% \begin{align*}
%% & \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
%% & \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
%% & \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
%% & \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
%% & \\
%% & \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
%% & \iterD \,f \,a = \iterD' \,f \,(f \,a)
%% \end{align*}
The definition above can be given the following intuitive
explanation. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
not terminate as well. If $f\,a$ terminates, there are two
possibilities: either $f\,a$ converges to $\inr\,b$, in which case
$\iterD\,f\,a$ terminates returning the value $b$; or $f\,a$ converges
to $\inl\,a'$, in which case we repeat the procedure by replacing $a$
with $a'$. Notice that in the latter case we also add one occurrence
of $\later$ to the total computation time. This addition is necessary
for ensuring the productivity of the corecursively defined function
$\iterD'$. In fact, by changing the second line of its specification to
$\iterD' \,f \,(\now \,(\inr\, a)) = \iterD' \,f \,(f a)$
and taking $f = \inrD$, we would have that $\iterD' \,f \,(\now
\,(\inr\, a))$ unfolds indefinitely without producing any output. In
Agda, such a definition would be rightfully rejected by the termination
checker.

@@

Terms of \Pio\ are modelled as morphism of $\Inv\Dapprox$,
i.e. partial isomorphisms. Here we only display the interpretation of a
selection of programs, we refer the interested reader to our Agda formalization for a
complete definition of the interpretation of terms.
\[
\begin{array}{ll}
\multicolumn{2}{l}{\semTm - : (A \lr B) \to \semTy A \, \rho \simeq
  \semTy B \, \rho}  \\
\semTm {f \oplus g} & = \semTm f +_{\D\simeq} \semTm g \\
\semTm {f \otimes g} & = \semTm f \times_{\D\simeq} \semTm g \\
\semTm {\trace\,f} & = \semTrace\,\semTm f
\end{array}
\]

Term equivalences of \Pio\ are modelled as morphism equalities in
$\Inv\Dapprox$, i.e. proofs of weak bisimilarity between two
morphisms. Formally, we define an operation:
\[
\semEq - : (f \LR g) \to \semTm f \approx \semTm g
\]
Again we refer the interested reader to our Agda formalization for a
complete definition of the interpretation of term equivalences.

\section{Conclusions}\label{sec:conclusions}
In this paper, we have extended the work of Carette and Sabry~\cite{CaretteS16}
to a (fully formalized) two-level calculus of \Pio{} programs and program
equivalences. Key in this effort was the use of the Kleisli category of the
delay monad on $\Set$ under weak bisimilarity, which turned out to support
iteration via a trace that preserves all partial isomorphisms, in this way
giving semantics to the dagger trace of \Pio. Further, the work was formalized
using Agda 2.6.0.

It is natural to wonder if our work can be ported to other monads of partiality
in Martin-L\"of type theory. As already discussed in the end of Section
\ref{sec:trace}, the maybe monad is not suitable for modelling a
well-behaved trace combinator without the assumption of classical principles
such as \textsf{LPO}. The partial map classifier \cite{UustaluV17-2,EscardoK17}
$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where $\mathsf{Prop}$ is
the type of propositions (types with at most one inhabitant), supports the
existence of a uniform iteration operator and therefore a trace. Nevertheless,
the specification of iteration is more complicated then the one presented in
Section \ref{sec:elgot} for the delay monad, which is a simple corecursive
definition. The complete Elgot monad structure of $\mathsf{PMC}$ follows from its
Kleisli category being a join restriction category, so iteration is defined in
terms of least upper bounds of certain chains of morphisms. The subcategory of
partial isomorphisms of the Kleisli category of $\mathsf{PMC}$ supports a
dagger trace combinator, which can be proved following the general strategy in
\cite{KaarsgaardAG17}. The exact same observations apply to the partiality
monad in homotopy type theory \cite{AltenkirchDK17,ChapmanUV19}, to which the
quotiented delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
isomorphic under the assumption of countable choice.
